BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-04-23 paper/paper.qmd
finish first pass of paper

@@

```julia
function multiscale_factorize(Y; kwargs...)
    scales, kwargs = initialize_scales(Y, kwargs)
    coarsest_scale, finer_scales... = scales

    # Factorize Y at the coarsest scale
    Yₛ = coarsen(Y, coarsest_scale; kwargs...)
    decomposition, stats, kwargs = factorize(Yₛ; kwargs...)

    # Factorize Y at progressively finer scales
    for scale in finer_scales
        # Use an interpolated version of the coarse factorization
        # as the initialization
        decomposition = interpolate(decomposition, scale; kwargs...)
        kwargs[:decomposition] = decomposition

        Yₛ = coarsen(Y, scale; kwargs...)
        decomposition, stats, kwargs = factorize(Yₛ; kwargs...)
    end
    return decomposition, stats, kwargs
end
```

@@

@@

# Conclusion
- all-in-one package
- provide a playground to invent new decompositions
- like auto-diff for factorizations

>>>>

Basic linear constrains `AX=B` are also implemented in the following manner. This constraint projects a factor `X` onto the affine space defined by `AX=B` for a linear operator or matrix `A` and bias `B`.

```julia
struct LinearConstraint{T <: Union{Function, AbstractArray}} <: AbstractConstraint
    linear_operator::T
    bias::AbstractArray
end

check(C::LinearConstraint{Function}, X::AbstractArray) = C.linear_operator(X) ≈ C.bias
check(C::LinearConstraint{<:AbstractArray}, X::AbstractArray) = C.linear_operator * X ≈ C.bias

# TODO implement linear constraint given an operator
function (C::LinearConstraint{Function})(X::AbstractArray)
    error("Linear Constraints defined in terms of an operator are not implemented (YET!)")
end

function (C::LinearConstraint{<:AbstractArray})(X::AbstractArray)
    error("Linear Constraints defined in terms of a general array are not implemented (YET!)")
end

function (C::LinearConstraint{<:AbstractMatrix})(X::AbstractArray)
    A = C.linear_operator
    b = C.bias
    X .-= A' * ( (A*A') \ (A*X .- b) ) # Projects X onto the subspace AX=b
end
```

@@

```julia
function multiscale_factorize(Y; kwargs...)
    continuous_dims, kwargs = initialize_continuous_dims(Y; kwargs...)
    scales, kwargs = initialize_scales(Y; kwargs...)
    coarsest_scale, finer_scales... = scales

    # Factorize Y at the coarsest scale
    Yₛ = coarsen(Y, coarsest_scale; dims=continuous_dims, kwargs...)

    constraints, kwargs = scale_constraints(Yₛ, coarsest_scale; kwargs...)
    decomposition, stats, _ = factorize(Yₛ; kwargs...)

    # Factorize Y at progressively finer scales
    for scale in finer_scales
        # Use an interpolated version of the coarse factorization
        # as the initialization.
        decomposition = interpolate(decomposition, 2; dims=continuous_dims, kwargs...)
        kwargs[:decomposition] = decomposition

        Yₛ = coarsen(Y, scale; dims=continuous_dims, kwargs...)

        constraints, kwargs = scale_constraints(Yₛ, scale; kwargs...)

        decomposition, stats, _ = factorize(Yₛ; kwargs...)
    end
    return decomposition, stats, kwargs
end
```

### Coarsening and Interpolating

@@

To apply a linear interpolation, we can first perform the constant interpolation and smooth out the result. The following code averages neighbouring values along the continuous dimensions specified. Since this is applied after the constant interpolation, every even indexed value becomes the average of its neighbours, and every odd indexed value remains fixed.

```julia
function linear_smooth!(Y, dims)
    all_dims = 1:ndims(Y)
    for d in dims
        axis = axes(Y, d)
        Y1 = @view Y[(i==d ? axis[begin+1:end-1] : (:) for i in all_dims)...]
        Y2 = @view Y[(i==d ? axis[begin+2:end] : (:) for i in all_dims)...]

        @. Y1 = 0.5 * (Y1 + Y2)
    end
    return Y
end
```

When interpolating an array that is an `AbstractDecomposition`, we can interpolate the factors directly instead of the combined array.

```julia
function interpolate(CPD::CPDecomposition, scale; dims=1:ndims(CPD), kwargs...)
    interpolated_matrix_factors = (d in dims ? interpolate(A, scale; dims=1, kwargs...) : A for (d, A) in enumerate(matrix_factors(CPD)))
    return CPDecomposition(Tuple(interpolated_matrix_factors))
end

function interpolate(T::Tucker1, scale; dims=1:ndims(T), kwargs...)
    core_dims = setdiff(dims, 1) # Want all dimensions except possibly the first
    interpolated_core = interpolate(core(T), scale; dims=core_dims, kwargs...)

    matrix = matrix_factor(T, 1)

    interpolated_matrix = 1 in dims ? interpolate(matrix, scale; dims=1, kwargs) : matrix
    return Tucker1((interpolated_core, interpolated_matrix))
end

function interpolate(T::Tucker, scale; dims=1:ndims(T), kwargs...)
    interpolated_matrix_factors = (d in dims ? interpolate(A, scale; dims=1, kwargs...) : A for (d, A) in enumerate(matrix_factors(T)))
    # Core is not interpolated
    return Tucker(Tuple(core(T), interpolated_matrix_factors...))
end
```

## Constraints with Multi-scale

Some constraints like `Entrywise` constrains can be used as-is at any scale, but other constraints like normalizations and linear constraints require some more thought. The main strategy is to modify constraints along continuous dimensions. Given a constraint on a factor, and the number of continuous dimensions it constrains, we can construct a modified constraint by scaling the full sized constraint appropriately.

For a $p$-norm constraint where some part of a factor $X_i$ needs to be normalized to $\lVert X_i \rVert_p=C$, we require the coarsened parts of the factor $\overline{X}_i$ to have norm $\lVert X_i \rVert_p= (C/s)^{1/p}$ where $s$ is the scale of the coarsening. For example, a scale of $3$ would mean we only take every $3$ entries of $X_i$.

For linear constraints `AX=B`, we construct a new constraint that coarsens $A$, and scales the bias $B$ by the scale. For example, if $x$ is a vector that represents a continuous function that is constrained to the affine space $Ax=b$, we treat the rows of the matrix $A$ as also being continuous functions that are inner product-ed with $x$ and can be coarsened in a similar manner to $x$. This means a `LinearConstraint(A, b)` gets scaled to the constraint `LinearConstraint(A[:, begin:scale:end], b ./ scale)`.

In all the constraints, the scale will need to be applied in each continuous dimension, so we implement this as `scale^n_continuous_dims`. The full implementation of `scale_constraint` is shown below.

```julia
function scale_constraint(constraint::AbstractConstraint, scale, n_continuous_dims)
    @warn "Unsure how to scale constraints of type $(typeof(constraint)). Leaving the constraint $constraint alone."
    return constraint
end

# Constraints that are fixed like entrywise constraints do not need to be scaled
function scale_constraint(constraint::Union{FIXED_CONSTRAINTS...}, scale, n_continuous_dims)
    return constraint
end

# TODO handle LinearConstraint{<:AbstractArray} or LinearConstraint{Function}
function scale_constraint(constraint::LinearConstraint{<:AbstractMatrix}, scale, n_continuous_dims)
    A = constraint.linear_operator
    b = constraint.bias
    return LinearConstraint(A[:, begin:scale:end], b ./ scale^n_continuous_dims)
end

function scale_constraint(constraint::ScaledNormalization{<:Union{Real,AbstractArray{<:Real}}}, scale, n_continuous_dims)
    norm = constraint.norm
    F = constraint.whats_normalized
    S = constraint.scale
    return ScaledNormalization(norm, F, S ./ scale^n_continuous_dims)
end

function scale_constraint(constraint::ScaledNormalization{<:Function}, scale, n_continuous_dims)
    norm = constraint.norm
    F = constraint.whats_normalized
    S = constraint.scale
    return ScaledNormalization(norm, F, (x -> x ./ scale^n_continuous_dims) ∘ S)
end

# TODO scale a projected normalization
function scale_constraint(constraint::ProjectedNormalization, scale, n_continuous_dims)
    @warn "Scaling ProjectedNormalization constraints is not implemented (YET!) Leaving the constraint $constraint alone."
    return constraint
end
```

We justify scaling constraints in this way with the following propositions and observations.

::: {#thm-rem-constraint-assumptions}

### Constraint Proposition Assumptions

In the following proposition, we assume $x\in\mathbb{R}^I$ represents an evenly-spaced discretization of an $L_f$-Lipschitz function $f:[l, u]\to\mathbb{R}$ on a finite interval $t\in[l, u]$ with entries

$$
x[i]=f(t[i]),
$$

$$
t[i] = l + (i-1)\Delta t= l + (i-1) \frac{u-l}{I-1}.
$$

We will use $\overline{x}\in\mathbb{R}^{\lfloor (I+1)/2 \rfloor}$ to represent the subvector made by removing every other entry of $x\in\mathbb{R}^I$ where

$$
\overline{x} = (x[1], x[3], x[5], \dots, x[\lfloor I \rfloor_{o}]).
$$

We use the shorthand

$$
\lfloor I \rfloor_{o} = 2\left\lfloor\frac{I+1}{2}\right\rfloor-1
$$

to round $I$ down to the nearest odd integer.

::: {#prp-linear-constraint-scaling}

### Linear Constraint Scaling

Let the assumption in @thm-rem-constraint-assumptions hold. Let $a\in\mathbb{R}^I$ be a discretization of an $L_g$-Lipschitz function $g:[l,u]\to\mathbb{R}$.

If $\langle a, x \rangle = b$, then for even $I$,

$$
\left\lvert \left\langle \overline{a}, \overline{x} \right\rangle - b/2 \right\rvert \leq \max(\lVert f \rVert_{\infty},\lVert g \rVert_{\infty}) \frac{I}{(I-1)} \frac{(L_{f} + L_{g})(u-l)}{4}
$$

and for odd $I$,

TODO check formatting
$$
\begin{align}
\left\lvert 2\left\lVert \overline{x} \right\rVert _1 - \frac{I+1}{I} \lVert x \rVert_{1} \right\rvert &= \left\lvert 2\left\lVert \overline{x} \right\rVert _1 - \lVert x \rVert_{1} - \frac{\lVert x \rVert_{1}}{I} \right\rvert \\
&=\left\lvert 2\sum_{i\ \text{odd}} \left\lvert x_{i} \right\rvert - \sum_{i=1}^I \left\lvert x_{i} \right\rvert  - \frac{\lVert x \rVert_{1}}{I} \right\rvert  \\
& = \left\lvert  \sum_{i\ \text{odd}}\left\lvert x_{i} \right\rvert+ \sum_{i\ \text{odd}}\left\lvert x_{i} \right\rvert - \left( \sum_{i\ \text{odd}}\left\lvert x_{i} \right\rvert + \sum_{i\ \text{even}} \left\lvert x_{i} \right\rvert  \right) - \frac{\lVert x \rVert_{1}}{I}  \right\rvert \\
& = \left\lvert  \sum_{i\ \text{odd}}\left\lvert x_{i} \right\rvert - \sum_{i\ \text{even}} \left\lvert x_{i} \right\rvert  - \frac{\lVert x \rVert_{1}}{I}  \right\rvert \\
&=  \left\lvert  \sum_{j=1}^{(I-1)/2}\left\lvert x_{2j-1} \right\rvert + \left\lvert x_{I} \right\rvert - \sum_{j=1}^{(I-1)/2} \left\lvert x_{2j} \right\rvert  - \frac{\lVert x \rVert_{1}}{I}  \right\rvert \\
&\leq \sum_{j=1}^{(I-1)/2}\left\lvert \left\lvert x_{2j-1} \right\rvert-\left\lvert x_{2j} \right\rvert \right\rvert + \left\lvert \left\lvert x_{I} \right\rvert - \frac{\lVert x \rVert_{1}}{I} \right\rvert \\
&\leq \sum_{j=1}^{(I-1)/2}\left\lvert  x_{2j-1} - x_{2j}  \right\rvert + \frac{1}{I} \left\lvert I \left\lvert x_{I} \right\rvert - \lVert x \rVert_{1} \right\rvert \\
&\leq \sum_{j=1}^{(I-1)/2} C + \frac{1}{I} \left\lvert \sum_{i=1}^I \left( \left\lvert x_{I} \right\rvert - \lvert x_{i} \rvert  \right)\right\rvert\\
&\leq C \frac{I-1}{2} + \frac{1}{I} \sum_{i=1}^I \left\lvert  \left\lvert x_{I} \right\rvert - \lvert x_{i} \rvert  \right\rvert\\
&\leq C \frac{I-1}{2} + \frac{1}{I} \sum_{i=1}^I \left\lvert   x_{I}  -  x_{i}  \right\rvert\\
&\leq C \frac{I-1}{2} + \frac{1}{I} \sum_{i=1}^I C (I-i)\ \text{(Apply Lipschitz recursively)} \\
& = C \frac{I-1}{2} + \frac{1}{I}\frac{CI(I-1)}{2}\\
&= \frac{L (u-l)}{2} + \frac{L (u-l)}{2} \\
&= L (u-l).
\end{align}
$$

::: {.proof}
See @sec-linear-constraint-scaling-proof.
:::

@prp-linear-constraint-scaling can be extended to a bound on the $L_1$ distance between $\overline{A}\overline{x}$ and $b/2$ where $Ax=b$ and $\overline{A}=A[:,begin:2:end]$ is the submatrix with every other column removed. TODO add corollary.

We also have a similar proposition for a $1$-norm constraint.

::: {#prp-l1-norm-constraint-scaling}

### $L_1$-norm Constraint Scaling

Let the assumption in @thm-rem-constraint-assumptions hold.

If $\lVert x \rVert_1 = b$, then for even $I$,

$$
\left\lvert \left\lVert \overline{x} \right\rVert_1 - b/2 \right\rvert \leq \frac{I}{(I-1)}\frac{L (u-l)}{4}
$$

and for odd $I$,

$$
\left\lvert \left\lVert \overline{x} \right\rVert_1 - \frac{I+1}{I} \frac{b}{2} \right\rvert \leq \frac{L(u-l)}{2}.
$$

::: {.proof}
See @sec-l1-norm-constraint-scaling-proof.
:::

Importantly, for normalized vectors $z = x / \lVert x \rVert_1$, the error bound in @prp-l1-norm-constraint-scaling (for even $I$) becomes

$$
\left\lvert \left\lVert \overline{z} \right\rVert_1 - 1/2 \right\rvert \leq \frac{I}{(I-1)}\frac{L (u-l)}{4b}.
$$

As the number of points get large, $I\to\infty$, a finer discretization will have more points and have a $1$-norm becoming unbounded $b\to\infty$. This means the error bound goes to zero and $\left\lVert\overline{z} \right\rVert_1 \to 1/2$.^[This also holds for odd $I$.]

@@

As implemented, both the regular and multi-scale approaches execute many iterations at the finest scale. The multi-scale method uses its coarsened iterations to warm start the fine iteration. So we should expect the multi-scale method to converge. We make this explicit with the following lemmas and theorems. The first subsection (@sec-analysis-lemmas) shows general functional and convex analysis facts about Lipschitz, smooth, and strongly convex functions, and what happens when you linearly interpolate Lipschitz functions. The next subsection (@sec-resolved-multiscale) shows convergence for the multi-scale method when we resolve the problem along the entire new discretization at each scale. And the final subsection (@sec-freezed-multiscale) looks at what happens when you freeze previously computed points so you only need to solve the problem at the interpolated points for each scale.

### Definitions

::: {#def-smooth-function}

A differentiable function $f:\mathbb{R}^I\to\mathbb{R}$ is $S$-smooth when $\nabla f$ is $S$-Lipschitz,

$$
\lVert \nabla f(x) - \nabla f(y) \rVert_2 \leq S \lVert x - y \rVert_2.
$$

::: {#def-strongly-convex-function}

A function $f:\mathbb{R}^I\to \mathbb{R}$ is $\mu$-strongly convex when $g(x) = f(x) - \frac{\mu}{2}\lVert x\rVert_2^2$ is convex.

When $f$ is differentiable, $f$ is $\mu$-strongly convex when

$$
f(y)\geq f(x)+\left\langle \nabla f(x),y-x  \right\rangle + \frac{\mu}{2}\left\lVert x-y \right\rVert_2^2
$$

for all $x,y\in \mathbb{R}^I$.

::: {#def-projected-gradient-decent}
One step of projected gradient descent for an $S$-smooth and $\mu$-strongly convex function $\mathcal{L}:\mathbb{R}^I\to\mathbb{R}$, a constraint set $\mathcal{C}\subseteq V$, at an iterate $x^k$ is the update,

$$
x^{k+1} \leftarrow P_{\mathcal{C}}\left( x^k - \frac{1}{S}\nabla\mathcal{L}(x^k) \right)
$$

where $P_{\mathcal{C}}:\mathbb{R}^I\to\mathcal{C}$ is the (Euclidean) projection operator onto $\mathcal{C}$.

::: {#prp-descent-lemma}

### Descent Lemma

In the projected gradient descent setting with an $S$-smooth and $\mu$-strongly convex function (@def-projected-gradient-decent), we can bound the distance to the unique minimizer $x^*\in\mathcal{C}$ in terms of the initial point $x^0\in\mathbb{R}^I$,

$$
\left\lVert x^{t}-x^* \right\rVert _2 \leq \left( 1-c \right)^t \left\lVert x^{0} -x^* \right\rVert _2
$$

with *condition number* $c=\mu/S$.

TODO cite this.

The vector space $\mathbb{R}^I$ in @def-strongly-convex-function, @def-smooth-function, and @def-projected-gradient-decent can be extended to tensor spaces $\mathbb{R}^{I_1 \times \dots \times I_N}$ with the Frobenius inner product $\langle X, Y\rangle _F$ and norm $\lVert \cdot \rVert _F$ more generally.

### Functional Analysis Lemmas {#sec-analysis-lemmas}

Before we can analyze the convergence of a multi-scaled method, we first need a handle on how linear interpolations play with discretized Lipschitz functions. To simplify the analysis of multi-scaled methods, we look at solving the general problem

$$
\min_{x \in \mathcal{C}} \mathcal{L} (x) := \sum_{i\in [I]} \ell_i(x[i])
$$

using projected gradient descent

$$
x \leftarrow P_\mathcal{C}\left(x - \frac{1}{L}\nabla \mathcal{L} (x)\right)
$$

where the samples $x[i]$ come from some $L_f$-Lipschitz function $f:\mathbb{R}\to\mathbb{R}$

$$
x[i] = f(t_i)
$$

on an interval $t\in [a,b]$ that has been evenly discretized $t_{i+1} - t_i=\Delta t$ for all $i\in[I]$.

We assume the loss functions $\ell_i$ over each entry $x[i]$ are $S$-smooth and $\mu$-strongly convex. A natural example of such a function $\mathcal{L}$ would be a least-squares loss

$$
\mathcal{L}(x) = \frac{1}{2} \left\lVert x - y \right\rVert _2^2 = \sum_{i\in[I]} \frac{1}{2} (x[i] - y[i])^2 = \sum_{i\in [I]} \ell_i (x[i]).
$$

The following lemmas explain that the smoothness and strong convexity of the functions $\ell_i$ carry over to the full loss $\mathcal{L}$.

::: {#lem-smooth-component-functions}

If $\ell_i:\mathbb{R}\to \mathbb{R}$ are each $S$-smooth, then $\mathcal{L}:\mathbb{R}^I\to \mathbb{R}$ is $S$-smooth.

::: {.proof}
First note that $(\nabla \mathcal{L}(x))_{i}= \frac{\partial}{\partial x[i]}\mathcal{L}(x)=\frac{\partial}{\partial x[i]}\sum_{j=1}^I\ell_i(x[j])=\sum_{j=1}^I\frac{\partial}{\partial x[i]}\ell_i(x[j])=\frac{\partial}{\partial x[i]}\ell_i(x[i])=\ell_i^'(x[i])$ since $\mathcal{L}$ is separable in each coordinate.
This gives us,
$$
\begin{align}
\left\lVert \nabla \mathcal{L}(x)-\nabla \mathcal{L}(y) \right\rVert_{2}^2 &= \sum_{i=1}^I(\nabla \mathcal{L}(x)-\nabla \mathcal{L}(y) )[i]^2 \\
&=\sum_{i=1}^I \left( \ell_i^'(x[i])-\ell_i^'(x[i]) \right)^2 \\
&\leq \sum_{i=1}^I S^2\left( x[i]-y[i] \right)^2 \tag{$\ell$ is $S$ smooth} \\
&=S^2\left\lVert x-y \right\rVert_{2}^2.
\end{align}
$$

Taking square roots completes the proof.

::: {#lem-strongly-convex-component-functions}

If $\ell_i:\mathbb{R}\to \mathbb{R}$ are all $m$ strongly convex, then $\mathcal{L}:\mathbb{R}^I\to \mathbb{R}$ is $m$ strongly convex (under the $2$-norm in $\mathbb{R}^I$).

Consider
$$
\begin{align}
\mathcal{L}(x)-\frac{m}{2}\lVert x \rVert_{2}^2 &= \sum_{i=1}^I\ell_i(x[i]) - \frac{m}{2}\sum_{i=1}^I x[i]^2 \\
&= \sum_{i=1}^I\left( \ell_i(x[i])- \frac{m}{2}x[i]^2 \right).
\end{align}
$$

The function $g(x[i])=\ell_i(x[i])- \frac{m}{2}x[i]^2$ is convex because each function $\ell_i(x[i])$ is $m$ strongly convex. The sum of convex functions is also convex, so $\mathcal{L}(x)-\frac{m}{2}\lVert x \rVert_{2}^2$ is convex.

These conditions can be relaxed to include larger sets of function like quasi-strongly convex functions [@necoara_linear_2019], but the general proof approach remains the same. We do require the separability of $\mathcal{L}$ so it can make sense to solve the problem over coarser discretizations.

Now we discuss how Lipschitz functions play with discretizations and linear interpolations.

::: {#lem-lipschitz-interpolation}

### Lipschitz Function Interpolation
Let $f:\mathbb{R}^n \to \mathbb{R}$, $a,b\in \mathbb{R}^n$, $t\in[0,1]$. Suppose $f$ is $L$-Lipshtiz. Then the error between the function at a point on the line segment between $a$ and $b$, and the linear interpolation is
$$
\left\lvert (tf(a)+(1-t)f(b)) - f(ta+(1-t)b) \right\rvert\leq 2Lt(1-t)\left\lVert a-b \right\rVert _2.
$$

Using the bound on the linear interpolation of a Lipschitz function repeatedly, for centre point interpolation ($t=1/2$ in @lem-lipschitz-interpolation), we can bound the error between an exact discretization of a function at a fine scale ($Y[j] = f(X[j])$) and a linear interpolation ($\hat{Y}$) coming from a coarser discretization ($y[k]=f(x[k])$).

::: {#lem-exact-interpolation}

### Exact Interpolation

Given $K$ uniformly space points $x[k]$ on $[a,b]$, (nearly) double them to get $J=2K-1$ uniformly spaced points $X[j]$ on $[a,b]$. Let $y[k]=f(x[k])$ for some $L$-Lipshitz function $f:\mathbb{R}\to \mathbb{R}$, and linearly interpolate the function values

$$
\hat{Y}[j]=\begin{cases}
y[\frac{j+1}{2}]   & \text{if $j$ is odd}\\
\frac{1}{2}\left( y[\frac{j}{2}] + y[\frac{j}{2}+1] \right) & \text{if $j$ is even}
\end{cases}\ ,
$$

where the true values are given by $Y[j]=f(X[j])$. Then the difference between the interpolated $\hat{Y}$ and exact values $Y$ is bounded by
$$
\left\lVert \hat{Y}-Y \right\rVert_{2}\leq \frac{L}{2\sqrt{ K-1 }}\left\lVert a-b \right\rVert_{2}.
$$

We also have an inexact version when we interpolate not from an exact coarse discretization ($y[k]=f(x[k])$), but from an approximate coarse discretization ($\tilde{y}[k] = f(x[k]) + \delta_k$).

::: {#lem-inexact-interpolation}

### Inexact Interpolation

Given a linear interpolation $\hat{\tilde{Y}}$ of an inexact discretization $\tilde{y}[k] = f(x[k]) + \delta_k$ of a function $f$ as described in @lem-exact-interpolation, where the interpolation is defined as

$$
\begin{align}
  \hat{\tilde{Y}}[j]&=\begin{cases}
  \tilde{y}[\frac{j+1}{2}]   & \text{if $j$ is odd}\\
  \frac{1}{2}\left( \tilde{y}[\frac{j}{2}] + \tilde{y}[\frac{j}{2}+1] \right) & \text{if $j$ is even}
  \end{cases},
\end{align}
$$

we have the error bound between the interpolated inexact discretization $\hat{\tilde{Y}}$ and the exact discretization of the function $Y$,

$$
\begin{align}
\left\lVert \hat{\tilde{Y}} - Y \right\rVert_2 &\leq \left\lVert \hat{\tilde{Y}} - \hat{Y} \right\rVert_2 + \left\lVert \hat{Y}-Y \right\rVert_2 \\
&\leq \sqrt{ 2 } \left\lVert \tilde{y}-y \right\rVert + \frac{L}{2\sqrt{ K-1 }}\left\lVert a-b \right\rVert_{2}.
\end{align}
$$

See @sec-inexact-interpolation-proof.

This makes sense in the following way: the error in our interpolation of approximate points, is bounded by two errors. The first comes from the fact that we interpolated using approximated values $\tilde{y}$ in place of the true values $y$, and the second comes from using a linear interpolation $\hat{Y}$ in place of the exact values $Y$.

### Re-solved Multi-scale {#sec-resolved-multiscale}

LEFT OFF HERE

We use the following general approach to show convergence.

Given some initial guess $x_S^0$ at the coarsest scale, we use the decent lemma (@prp-descent-lemma) to bound the error between our iterate $x_S^{K_S}$ after $K_S$ iterations, and the solution $x_S^*$ at the scale $S$. We can linearly interpolate our point $\hat{x}_S^{K_S}$ and use it to initialize another round of projected gradient descent $x_{S-1}^0=\hat{x}_S^{K_S}$ at the slightly finer scale. We can link the error between our iterate $x_S^{K_S}$ and the solution $x_S^*$ at  scale $S$ with the initial error between an iterate at the finer scale $x_{S-1}^0$ and the solution $x_{S-1}^*$ at this scale using the inexact interpolation lemma (@lem-inexact-interpolation). We perform $K_s$ iterations at each scale $s=S,S-1,\dots,2,1$ to get an error bound between our iterate at the finest scale $x_1^K_1$ the solution at this scale $x_1^*$ in terms of the initial error at the coarsest scale $x_S^0$.

We present the main descent theorem here, and leave the rest of the details in the appendix.

::: {#thm-resolved-multiscale-descent-error}

### Re-solved Multi-scale Descent Error

Let $c$ be the condition number of the loss function $\mathcal{L}$, $L_f$ be the Lipschitz constant for the underlying continuous function $f$, and $S$ be the coarsest scale. For the setting described in @sec-analysis-lemmas, we have the following error bound on the fixed multi-scaled method.
$$
\begin{align}
&\left\lVert x^0_{1} - x^*_{1} \right\rVert _2 \\
&\leq \sqrt{ 2 ^{S-1}}\left(1-c\right)^{\sum_{s=1}^{S} K_{s}}\left\lVert x_{S}^{0} -x_{S}^* \right\rVert _2 + \frac{L_f\left\lvert a-b \right\rvert}{2\sqrt{2^{S+1}}}\sum_{s=1}^{S-1}  2^{s} \left(1-c\right)^{\sum_{t=1}^sK_{t}}
\end{align}
$$

See @sec-resolved-multiscale-descent-error-proof

### Freezed Multi-scale {#sec-freezed-multiscale}

We take a nearly identical approach to the re-solved multi-scale method as described in @sec-resolved-multiscale, but we instead freeze the entries of our interpolation $\hat{x}_s^{K_s}$ that correspond to the same points as the prior iteration $x_s^{K_s}$. The projected gradient update will only act on the interpolated and unfrozen values which we will call $x_{(s)}^k$.

TODO can I just introduce this notation in the proofs so that the main paper is cleaner?

To be clear let us look at an example where the finest scale has $2^3+1=9$ points. After performing $K_3$ iterations at the coarsest scale $S=3$ with a total of three points, and slightly finer scale $S=2$ with five points (but only two unfrozen points), the full vector at the finest scale would be

$$
\begin{align}
& x_{1}^k = \\
& \left( {{x}_{(3)}^{K_{3}}}[1],{ x_{(1)}^{k}}[1],{ x_{(2)}^{K_{2}}}[1], { x_{(1)}^{k}}[2], {{x}_{(3)}^{K_{3}}}[2], { x_{(1)}^{k}}[3], { x_{(2)}^{K_{2}}}[2], { x_{(1)}^{k}}[4], {{x}_{(3)}^{K_{3}}}[3] \right)
\end{align}
$$

where the vector of free variables is

$$
\begin{align}
& x_{(1)}^k = \left( { x_{(1)}^{k}}[1], { x_{(1)}^{k}}[2], { x_{(1)}^{k}}[3], { x_{(1)}^{k}}[4]\right).
\end{align}
$$

**Summary of notation**

- $\hat{x}$ is some approximation of $x$
- $x_{(s)}$ is a vector of just the free variables at scale $s$
- $x_{s}$ is a vector of the free and fixed variables
- $x^k$ is the $k$th iteration
- $K_{s}$ is the number of iterations performed at scale $s$
- $e=\hat{x}-x^*$ is the error
- have $e_{(s)},e_{s},e^k$ similarly.
	- Example, $e_{1}^2=x_{1}^2-x_{1}^*$ is the error between our iterates at the finest scale $s=1$ and the true values $x_{1}^*$ after $2$ iterations

We present the main descent theorem here, and leave the rest of the details in the appendix.

::: {#thm-freezed-multiscale-descent-error}

### Freezed Multi-scale Descent Error

Let $c$ be the condition number of the loss function $\mathcal{L}$. For the setting described in @sec-analysis-lemmas, we have the following error bound on the freezed multi-scaled method.

$$
\begin{align}
\left\lVert e_{1}^{K_{1}} \right\rVert &\leq \left\lVert e_{S}^{0}  \right\rVert (1-c)^{K_{S}}\prod_{s=1}^{S-1}(1+(1-c)^{K_{s}}) \\
&\qquad+\ \frac{L_f}{2}\left\lvert b-a \right\rvert\sum_{s=1}^{S-1} \frac{1}{\sqrt{ 2^{S-s} }}(1-c)^{K_{s}}\prod_{j=1}^{s-1}\left( 1+(1-c)^{K_{j}} \right).
\end{align}
$$

If we use the same number of iterations $K_s=K$ at each scale, this reduces to the closed form upper bound
$$
\begin{align}
\left\lVert e_{1}^{K_{1}} \right\rVert &\leq \left\lVert e_{S}^{0}  \right\rVert d(K)(1+d(K))^{S-1}  \\
&\qquad+\ \frac{L_f}{2}\left\lvert b-a \right\rvert\frac{d(K) \left( \sqrt{2^S} \left( d(K) + 1 \right)^S - \sqrt{2} \left( d(K) + 1 \right) \right)}{\sqrt{2^S} \left( d(K) + 1 \right) \left( \sqrt{2} d(K) + \sqrt{2} - 1 \right)}
\end{align}
$$

where $d(K) = (1-c)^K$ is a small number between $0 < d(K) < 1$.

See @sec-freezed-multiscale-descent-error-proof.

We summarize the convergence with @cor-multiscale-convergence that sends the number of iterations to infinity.

::: {#cor-multiscale-convergence}

### Multiscale Convergence

As the total number of iterations grows $\sum_{s=1}^S K_s \rightarrow \infty$ (in the case of re-solve multi-scale in @thm-resolved-multiscale-descent-error) or the number of iterations at each scale $K_s\to \infty$ (in the case of freezed multi-scale in @thm-freezed-multiscale-descent-error), the final error goes to $0$, $\lVert e_1^{K_1} \rVert_2 \rightarrow 0$, and we converge to the solution $x_1^{K_1} \rightarrow x_1^*$ at the finest scale.

## Comparison With Projected Gradient Descent

From @prp-descent-lemma, projected gradient descent with a generic initialization on the fine grid gives us the following iterate convergence

$$
\left\lVert x_{1}^{K}-x_{1}^* \right\rVert\leq \left( 1-c \right)^{K} \left\lVert x_{1}^{0} -x_{1}^* \right\rVert.
$$

We can use this in combination with an expected initial error to get an expected number of iterations needed until we have converged to a desired tolerance. This is make precise by @thm-expected-pgd-convergence.

::: {#thm-expected-pgd-convergence}

### Expected Projected Gradience Descent Convergence

Assume the problem is either scaled or shifted so that the solution is normalized $\left\lVert x_{1}^* \right\rVert _2 = 1$ or centred $\left\lVert x_{1}^* \right\rVert _2 = 0$ where $x_{1}^*\in\mathbb{R}^{I}$. Choose an initialization $x_{1}^0 \in \mathbb{R}^I$ with i.i.d. standard normal entries $(x^0_{1})_{i}\sim \mathcal{N}(0,1)$. Then the expected initial error is
$$
\mathbb{E} \left\lVert x_{1}^{0} -x_{1}^* \right\rVert_2 =\sqrt{ I+1 }.
$$

And, as the number of points grows $I\to \infty$, if we iterate projected gradient descent $K$ times where
$$
 K \geq \frac{\log(1/\epsilon)+{\log(I-1)}/{2}}{-\log \left( 1-c \right)},
$$

the expected error is less than $\epsilon$,

$$
\mathbb{E} \left\lVert x_{1}^{0} -x_{1}^* \right\rVert_2 \leq \epsilon.
$$

See @sec-expected-pgd-convergence-proof.

This is contrasted with the expected initial and final error for re-solved multi-scaled descent (@thm-expected-resolved-convergence) and freezed multi-scaled descent (@thm-expected-freezed-convergence).

::: {#thm-expected-resolved-convergence}

### Expected Re-solved Multi-scale Convergence

Assume the same setting as in @thm-expected-pgd-convergence, but instead using re-solved multi-scaled descent method starting with a scale $s_0$. Let the number of points at the finest scale be $I = 2^S + 1$. Then we have the expected initial error

$$
\mathbb{E} \left\lVert x_{s_{0}}^{0} -x_{s_{0}}^* \right\rVert_2 = \sqrt{ 2^{S-s_{0}+1} + 2 }.
$$

Performing $K_s$ iterations at each scale gives us the expected final error

$$
\begin{align}
&\mathbb{E} \left\lVert x_{1}^{K_1} -x_{1}^* \right\rVert_2 \\
&\leq (1-c)^{K_{1}}\sqrt{ 2 ^{S+1}}\left((1-c)^{\sum_{s=2}^S K_{s}}+ \frac{L\left\lvert a-b \right\rvert}{2\cdot{2^{S}}}+ \frac{L\left\lvert a-b \right\rvert}{2\cdot {2}}\sum_{s=2}^{S-1}  \frac{(1-c)^{\sum_{t=2}^sK_{t}}}{2^{S-s}} \right).
\end{align}
$$

See @sec-expected-resolved-convergence-proof.

It is not as straightforward to get the required number of iterations to achieve a desired level of accuracy. Additionally, this would not be a fair comparison with projected gradient descent since we expect an iteration at a coarse scale $S$ to be cheaper (less time and fewer floating point operations) than an iteration at the finest scale $s=1$. For this reason, we need to cost an iteration of projected gradient descent at a scale $s$ in terms of the size of the problem at that scale.

::: {#lem-cost-of-gd-vs-ms}

### Cost of Projected Gradient Descent vs Multi-scale

Suppose the total cost of regular descent is given by
$$
C_{\text{GD}}=C_{1}K
$$

where $C_1$ is the cost of performing one iteration at the finest scale $s=1$.

The total cost of multi-scale descent is
$$
C_{\text{MS}} = \sum_{s=1}^S C_{s}K_{s}
$$

similarly.

If we assume the cost of projected gradient scales at least in the size of the problem ($C_{1} = \Omega(I)$, i.e. $C_1 \geq CI$ for some $C\geq 0$), then the cost of projected gradient descent at scale $s$ is
$$
C_{s} \geq \frac{ 2^{S-s+1}+1 }{ 2^{S-s}+1}C_{s+1} \geq \frac{3}{2} C_{s+1},
$$

which gives the total cost of multi-scale descent at most
